Efficient pedigree recording for fast population genetics simulation

### 2018-01-12 forwards_paper.tex
Updated description of alg S.

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\fwdpy}{\texttt{fwdpy11}}
\newcommand{\justc}{\texttt{C}}
\newcommand{\cpp}{\texttt{C++}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\ftprime}{\texttt{ftprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
\newcommand{\priorityq}[1]{\mbox{\textbf{PriorityQueue}}({#1})}
\newcommand{\listnew}[1]{\mbox{\textbf{List}}(#1)}
\newcommand{\pqpush}[2]{#1.\mbox{\textbf{push}}(#2)}
\newcommand{\pqpop}[1]{#1.\mbox{\textbf{pop}}()}
\newcommand{\listappend}[2]{#1.\mbox{\textbf{append}}(#2)}
\newcommand{\ancsegment}[1]{\mbox{\textbf{Segment}}(#1)}
\newcommand{\taddrow}[2]{#1.\mbox{\textbf{addrow}}(#2)}
\newcommand{\attrparent}[1]{#1.\mbox{parent}}
\newcommand{\attrchild}[1]{#1.\mbox{child}}
\newcommand{\attrleft}[1]{#1.\mbox{left}}
\newcommand{\attrright}[1]{#1.\mbox{right}}
\newcommand{\attrid}[1]{#1.\mbox{id}}
% mbox to avoid hyphenating
\newcommand{\tsimplify}[1]{\mbox{\textbf{simplify}}(#1)}
\newcommand{\tsort}[1]{\mbox{\textbf{sort}}(#1)}
\newcommand{\tnodetable}[1]{\mbox{\textbf{NodeTable}}(#1)}
\newcommand{\tedgetable}[1]{\mbox{\textbf{EdgeTable}}(#1)}

@@

\plr{Replace this with Jerome's interval-tree implementation?}

Here we describe the simplification algorithm.
The flow is close to our implementation,
but the description below uses set operations,
while in the implementation,
all ancestors are maintained as linked lists of segments
ordered by left endpoint.
As it goes, the algorithm
records a list $M$ that gives the mapping
from input nodes to output nodes,
so that if $M[p] = u$ then the $p^\text{th}$ node in the input
is the same as the $u^\text{th}$ node in the output.
The current state of the algorithm is recorded in $\Al$,
a list of ancestors
each of which is a collection of ancestral segments.
An ancestral segment is a triple $(\ell, r, x)$,
indicating that the output node $x$ inherits from this ancestor
on the genomic segment $[\ell, r)$.

@@

\algstep{S0.}{Initialize.}{
    Set $\Al_u \leftarrow \listnew{}$ for $0 \leq u < |\Nt_I|$. Then,
    for each $u$ in $S$, set $v \leftarrow \taddrow{\Nt_O}{\Nt_I[u]}$ and
    call $\listappend{\Al[u]}{\ancsegment{0, L, v}}$. Then set $Q \leftarrow \priorityq{}$
    and $u \leftarrow 0$.
}

@@

\algstep{S2.}{Insert edge/ancestry intersections.}{
    For each $e\in E$ and each $x \in \Al[\attrchild{e}]$,
    if $\attrright{x} > \attrleft{e}$ and $\attrright{e} > \attrleft{x}$,
    set $\ell \leftarrow \max(\attrleft{x}, \attrright{e})$,
    $r \leftarrow \min(\attrright{x}, \attrright{e})$,
    $y \leftarrow \ancsegment{\ell, r, \attrid{x}}$ and call
    $\pqpush{Q}{y}$.
}

@@

\algstep{S4.}{No overlapping segments.}{
    Set $x \leftarrow X[0]$ and $\alpha \leftarrow x$. If $|Q| > 0$ and
    $\attrleft{Q[0]} < \attrright{x}$, set $\alpha \leftarrow
    \ancsegment{\attrleft{x}, \attrright{Q[0]}, \attrid{x}}$,
    $\attrleft{x} \leftarrow \attrleft{Q[0]}$ and call $\pqpush{Q}{x}$.
    Go to \algref{S7}.
}

@@

\algstep{S6.}{Record edges.}{
    Set $\alpha \leftarrow \ancsegment{\ell, r, v}$. Then for each $x \in X$
    call $\taddrow{\Et_O}{l, r, v, \attrid{x}}$ and, if $\attrright{x} > r$,
    set $\attrleft{x} \leftarrow r$ and call $\pqpush{Q}{x}$.
}

\algstep{S7.}{Left coordinate loop.}{
    Call $\listappend{\Al}{\alpha}$. If $|Q| > 0$ return to \algref{S3}.
}

@@

% The algorithm steps back up through the pedigree,
% since nodes are ordered by time-since-birth.
% First, for each input node $p$, we must find all ancestral segments inheriting from $p$ --
% so, for every edge $(\ell, r, p, c)$,
% Algorithm \algref{R} replaces any ancestry segment $[u, v)$ held by $c$
% with $[u,v) \setdiff [\ell, r)$
% (deleting the segment or splitting in two if necessary),
% and adding the removed segment $[u,v) \cap [\ell, r)$ to the merge queue $Q$:

% Next we must merge the ancestry segments in $Q$,
% creating the new ancestor corresponding to input node $p$,
% adding a node and edges to the output if any segments overlap,
% which implies that an output coalescence has occurred.
% Below we index elements of an ancestral segment $z = (\ell, r, u)$
% by writing $z.\lleft = \ell$ and $z.\lright = r$ for the endpoints,
% and $z.\lid = u$ for the output node ID.

>>>>

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\fwdpy}{\texttt{fwdpy11}}
\newcommand{\justc}{\texttt{C}}
\newcommand{\cpp}{\texttt{C++}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\ftprime}{\texttt{ftprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
\newcommand{\priorityq}[1]{\mbox{\textbf{PriorityQueue}}({#1})}
\newcommand{\listnew}[1]{\mbox{\textbf{List}}(#1)}
\newcommand{\pqpush}[2]{#1.\mbox{\textbf{push}}(#2)}
\newcommand{\pqpop}[1]{#1.\mbox{\textbf{pop}}()}
\newcommand{\listappend}[2]{#1.\mbox{\textbf{append}}(#2)}
\newcommand{\ancsegment}[1]{\mbox{\textbf{Segment}}(#1)}
\newcommand{\taddrow}[2]{#1.\mbox{\textbf{addrow}}(#2)}
\newcommand{\attrparent}[1]{#1.\mbox{parent}}
\newcommand{\attrchild}[1]{#1.\mbox{child}}
\newcommand{\attrleft}[1]{#1.\mbox{left}}
\newcommand{\attrright}[1]{#1.\mbox{right}}
\newcommand{\attrnode}[1]{#1.\mbox{node}}
% mbox to avoid hyphenating
\newcommand{\tsimplify}[1]{\mbox{\textbf{simplify}}(#1)}
\newcommand{\tsort}[1]{\mbox{\textbf{sort}}(#1)}
\newcommand{\tnodetable}[1]{\mbox{\textbf{NodeTable}}(#1)}
\newcommand{\tedgetable}[1]{\mbox{\textbf{EdgeTable}}(#1)}

@@

Here we describe the simplification algorithm. The algorithm uses a few simple structures
besides the node and edge tables discussed earlier. As the algorithm needs to track the mapping
between input and output node IDs over specific genomic intervals, we use a $\ancsegment$
type to track these individual mappings. Thus, a segment $x$ has three attributes:
$\attrleft{x}$ and $\attrright{x}$ represents the genomic interval in the usual way,
and $\attrnode{x}$ represents the output node ID that is assigned to this interval. The per-interval
segments are assigned to input IDs via the $\Al$ mapping, which is the main state required
by the algorithm. Each element $\Al[u]$
is a list of ancestral segments, describing the output node ID mapped to input
node ID $u$ over a set of intervals. For each input node, we need to find the process
the resulting set of ancestral segments to find overlaps (and hence coalescences). We
do this using a $\priorityq{}$ type. This priority queue maintains a list of ancestral
segments in sorted order, such that the segements with the smallest left coordinate
are at the start of the list. Given a priority queue $Q$, the operation
$\pqpush{Q}{x}$ inserts the segment $x$ into the queue, and $\pqpop{Q}{x}$
returns (and removes) the segment in $Q$ with the smallest left coordinate.
To obtain the segment currently in the queue with the smallest left value we
write $Q[0]$.

A step-by-step explanation of the algorithm follows the algorithm listing.

@@

\algstep{S0.}{Initialisation.}{
    Set $\Al_u \leftarrow \listnew{}$ for $0 \leq u < |\Nt_I|$. Then,
    for each $u$ in $S$, set $v \leftarrow \taddrow{\Nt_O}{\Nt_I[u]}$ and
    call $\listappend{\Al[u]}{\ancsegment{0, L, v}}$. Then set $Q \leftarrow \priorityq{}$
    and $u \leftarrow 0$.
}

@@

\algstep{S2.}{Insert edge/ancestry intersections.}{
    For each $e\in E$ and each $x \in \Al[\attrchild{e}]$,
    if $\attrright{x} > \attrleft{e}$ and $\attrright{e} > \attrleft{x}$,
    set $\ell \leftarrow \max(\attrleft{x}, \attrright{e})$,
    $r \leftarrow \min(\attrright{x}, \attrright{e})$,
    $y \leftarrow \ancsegment{\ell, r, \attrnode{x}}$ and call
    $\pqpush{Q}{y}$.
}

@@

\algstep{S4.}{No overlapping segments.}{
    Set $x \leftarrow X[0]$ and $\alpha \leftarrow x$. If $|Q| > 0$ and
    $\attrleft{Q[0]} < \attrright{x}$, set $\alpha \leftarrow
    \ancsegment{\attrleft{x}, \attrright{Q[0]}, \attrnode{x}}$,
    $\attrleft{x} \leftarrow \attrleft{Q[0]}$ and call $\pqpush{Q}{x}$.
    Go to \algref{S7}.
}

@@

\algstep{S6.}{Record edges.}{
    Set $\alpha \leftarrow \ancsegment{\ell, r, v}$. Then for each $x \in X$
    call $\taddrow{\Et_O}{l, r, v, \attrnode{x}}$ and, if $\attrright{x} > r$,
    set $\attrleft{x} \leftarrow r$ and call $\pqpush{Q}{x}$.
}

\algstep{S7.}{Left coordinate loop.}{
    Call $\listappend{\Al[u]}{\alpha}$. If $|Q| > 0$ return to \algref{S3}.
}

@@

The algorithm begins in \algref{S1} by allocating an empty list to store ancestral
segments for each of the $|\Nt_I|$ input nodes, and then creating the initial
state for the samples in $S$. For each input sample with node ID $u$, we
add a row in the output node table $\Nt_O$, and create a mapping for this
new node $v$. After this initial state has been created, we then allocate our
priority queue $Q$, and set the current input ID $u$ to $0$.

The algorithm steps back up through the pedigree and steps \algref{S1}--\algref{S8}
consider each node $u$ in turn (note we are assuming that nodes are ordered by time-since-birth).
In \algref{S1} we first find all edges in the input where the parent is equal to
our current input ID $u$, and then set the corresponding output ID $v$ to $-1$.
Then, in \algref{S2} we find all intersections between these edges and the
mapped ancestry segments for their child nodes. For each of these intersections we
create a new segment $y$ and insert it into the priority queue.

After finding all overlaps and filling the priority queue with the resulting ancestry
segments, we then need to process these segments. Steps \algref{S3}--\algref{S7}
loop over the segments in the queue, considering each in turn until the queue is empty.
Step \algref{S3} builds a list $X$ of all the segments that have the
minimum left coordinate (along with some bookkeeping to keep track of where our next
right coordinate $r$ is), removing them from $Q$. If there is only one element in $X$
this means that we have no overlapping ancestral segments at this coordinate and
we proceed to \algref{S4}.

As no overlaps occur, there is no change in the mapping between input and output IDs
and so the ancestry segment $\alpha$ is not changed (this is assigned to $\Al_u$ in \algref{S7}).
However, there is a complication caused by other segments which may intersect with $x$,
which have left coordinate greater then $\attrleft{x}$. If $x$ intersects with this value,
then there are other overlaps we must consider. Therefore, we change the left coordinate
of $x$ to this coordinate and reinsert it into $Q$ where it can be processed again
later. After doing this, we proceed to \algref{S7}, update the ancestry mapping for
the input node $u$, and loop back to \algref{S3} if $Q$ is not empty.

On the other hand, if there is more than one segment in $X$, we know that an
overlap among the ancestral segments has occured and we must update the output
nodes and edges accordingly. In step \algref{S5} we first allocate a new output
node $v$ for $u$, if it has not been done already (we may have many overlapping
segments along the genome). Step \algref{S6} then continues by creating a new
ancestry segment $\alpha$ mapping the current interval to the output node $v$.
We then iterate over all segments in $X$, adding the appropriate output edges
and editing and reinserting the ancestry segment $x$ into the priority queue if
necessary, as before.
